Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-10-17 analysis/paper/London_HauloutBeahvior.Rmd
solar radiation supplemental + fits

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(gt)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)
library(MetBrewer)

@@

deploy_table %>%
  dplyr::group_by(species, sex, age) %>%
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>%
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>%
  dplyr::select(-c("n","seal_hours")) %>%
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>%
  flextable() %>%
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>%
  merge_h(part = "header") %>%
  align(align = "left", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  fontsize(size = 9, part = "all") %>% 
  width(width = 1.35) %>% 
  width(j = ~ Species, width = 0.9) %>% 
  width(j = ~ Sex, width = 0.4)

@@

covariate_tbl %>% 
  flextable() %>% 
  fontsize(size = 9, part = "all") %>% 
  hline(i=8) %>% 
  width(width = 1.0) %>% 
  width(j = ~ Description, width = 2.5)
```

@@

  deploy_tbl <- as_grouped_data(deploy_tbl,groups="species") %>% 
  flextable::as_flextable(hide_grouplabel = TRUE) %>% 
    width(j = c('agency','deploy_year','location','publications','permits',
                'n_seals', 'sex', 'age_class'), 
          width = c(1, 0.8, 1.5, 1.5, 0.8, 0.8, 0.8, 1.5)) %>%
    align(j = 'n_seals', align = 'left', part = 'all') %>%
  set_header_labels(
    agency = "Institution",
    deploy_year = "Year Deployed",
    location = "Location",
    publications = "Publication(s)",
    permits = "Permits",
    n_seals = "No. Seals",
    sex = "Sex",
    age_class = "Age_Class"
  ) %>%
  bold(j = 1, i = ~ !is.na(species), bold = TRUE, part = "body") %>%
  bold(part = "header", bold = TRUE) %>%
  add_footer_lines(as_paragraph(
    as_b("ADFG"),"=Alaska Department of Fish and Game; ", as_b("NSB"),"=North Slope Borough; ", as_b("NMFS"),"=NOAA National Marine Fisheries Service; ", as_b("NPWC"),"=North Pacific Wildlife Consortium"
  )) %>%
    fontsize(size = 10, part ="header") %>%
    fontsize(size = 9, part = "body") %>%
    fontsize(size = 8, part = "footer") %>%
  set_table_properties(layout = "fixed",
                       opts_pdf = list('caption_repeat' = FALSE))

@@

>>>>

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(gt)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)
library(MetBrewer)
library(terra)

@@

deploy_table %>%
  dplyr::group_by(species, sex, age) %>%
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>%
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>%
  dplyr::select(-c("n","seal_hours")) %>%
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>%
  flextable() %>%
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>%
  merge_h(part = "header") %>%
  flextable::align(align = "left", part = "header") %>% 
  flextable::align(i = 1, align = "center", part = "header") %>% 
  fontsize(size = 9, part = "all") %>% 
  flextable::width(width = 1.35) %>% 
  flextable::width(j = ~ Species, width = 0.9) %>% 
  flextable::width(j = ~ Sex, width = 0.4)

@@

covariate_tbl %>% 
  flextable() %>% 
  fontsize(size = 9, part = "all") %>% 
  hline(i=8) %>% 
  flextable::width(width = 1.0) %>% 
  flextable::width(j = ~ Description, width = 2.5)
```

@@

  deploy_tbl <- as_grouped_data(deploy_tbl,groups="species") %>% 
  flextable::as_flextable(hide_grouplabel = TRUE) %>% 
    flextable::width(j = c('agency','deploy_year','location','publications','permits',
                'n_seals', 'sex', 'age_class'), 
          width = c(1, 0.8, 1.5, 1.5, 0.8, 0.8, 0.8, 1.5)) %>%
    flextable::align(j = 'n_seals', align = 'left', part = 'all') %>%
  set_header_labels(
    agency = "Institution",
    deploy_year = "Year Deployed",
    location = "Location",
    publications = "Publication(s)",
    permits = "Permits",
    n_seals = "No. Seals",
    sex = "Sex",
    age_class = "Age_Class"
  ) %>%
  bold(j = 1, i = ~ !is.na(species), bold = TRUE, part = "body") %>%
  bold(part = "header", bold = TRUE) %>%
  add_footer_lines(as_paragraph(
    as_b("ADFG"),"=Alaska Department of Fish and Game; ", as_b("NSB"),"=North Slope Borough; ", as_b("NMFS"),"=NOAA National Marine Fisheries Service; ", as_b("NPWC"),"=North Pacific Wildlife Consortium"
  )) %>%
    fontsize(size = 10, part ="header") %>%
    fontsize(size = 9, part = "body") %>%
    fontsize(size = 8, part = "footer") %>%
  set_table_properties(layout = "fixed",
                       opts_pdf = list('caption_repeat' = FALSE))

@@

## Exploring Insolation (Solar Radiation) as a Model Covariate

### Introduction

During the peer review process for this manuscript, Anthony Fishbach suggested
the possibility of using predicted insolation (or solar radiation) values from
the reanalysis model as a more direct and, potentially, more informative
predictor of the daily haul-out cycle in seals compared to time of day. The
notion being that seals are directly responding to changes in solar radiation
throughout the day and not what time of day it is (i.e. seals don't have
watches). Additionally, given the energetic benefits of increased solar
radiation it could be more informative as we would expect seals might have a
higher haul-out probability on sunnier days and for there to be geographic
variability in haul-out behavior associated with geographical differences in
insolation. This approach has an additional benefit of being more parsimonious
compared to our use of the Fourier series or other approaches to represent
hour-of-day in the model (e.g. 24 factors for each hour).

Because of these reasons, we spent some time considering and exploring this
possibility for our model and the analysis presented in this manuscript. A key
drawback to reliance on solar insolation, in our minds, is that we would lose
insight regarding potential crepuscular patterns -- insolation does not 
differentiate between dusk or dawn. Bi-modal patterns have been
previously observed in ringed seals and our results in this analysis show some indication of
increased haul-out probability during dawn and dusk periods for bearded seals.
For other phocid speices, increased haul-out probability before solar noon or after
solar noon has been observed. Importantly, understanding these relationships
between haul-out probability and hour-of-day can have important ramifications
on aerial survey study design -- a key focus of this paper.  
Another hesitation we had was that insolation estimates from reanalysis models have
not been previously used as a model covariate within a published study of 
pinniped haul-out behavior. Thus, for this analysis, we chose to keep our
original approach and rely on the Fourier series to capture any hour-of-day
effects.

That said, we think the idea of insolation as a model covariate in pinniped
haul-out models is intriguing and worth further exploration. The current
availability and increased accessibility to detailed climate reanalysis
products that include insolation is exciting and we encrourage future, more
detailed exploration of this as a component in pinniped haul-out analysis. To
provide some inspiration, we present some initial efforts and model 
comparisons.

### Methods

In this manuscript, we rely on the NARR reanalysis model as the source for our
weather covariates. However, since our initiation of this analysis, the ERA5
reanalysis model (https://doi.org/10.24381/cds.adbb2d47) has become one of the
go-to standards for global climate reanalysis and provides an increased temporal
resolution to hourly (compared to the 3-hour resolution of NARR). The global
coverage of ERA5 provides additional flexibility in that the area of interest is
not limited to North America. The ERA5 model provides a number of solar
radiation parameters and it is important to evaluate and understand each of
these estimates in order to select the one that is likely most relevant to
seals. Here, we use the 'surface short-wave (solar) radiation downwards.' This
parameter is described as _the amount of solar radiation (also known as shortwave radiation)
that reaches a horizontal plane at the surface of the Earth and comprises both
direct and diffuse solar radiation. To a reasonably good approximation, this
parameter is the model equivalent of what would be measured by a pyranometer (an
instrument used for measuring solar radiation) at the surface_ 
(https://codes.ecmwf.int/grib/param-db/?id=169). Thus, this is the
parameter which most closely represents the amount of solar radiation likely
felt by a seal hauled out of the water.

ERA5 data is available via the Copernicus climate data store API which can be
queried with the CDS-API Python package 
(https://cds.climate.copernicus.eu/api-how-to). The R code provided here
documents the download of the _surface_solar_radiation_downwards_ parameter for
our study area of interest and years of interest. The `reticulate` R package
(https://CRAN.R-project.org/package=reticulate) allows interaction with Python.
Additionally, note, extra steps are required to download data on either side
of the 180 anti-meridian.

```{r cdsapi, eval=FALSE, include=TRUE, echo = TRUE}
library(tidyverse)
library(reticulate)
library(sf)
library(terra)

#import python CDS-API
cdsapi <- import('cdsapi')
#for this step there must exist the file .cdsapirc
server = cdsapi$Client() #start the connection

get_era5 <- function(y) {
  #we create the query
  query <- r_to_py(
    list(
      variable = "surface_solar_radiation_downwards",
      product_type = "reanalysis",
      area = "75/152/47/180", # North, West, South, East
      year = y,
      month = str_pad(2:7, 2, "left", "0"),
      day = str_pad(1:31, 2, "left", "0"),
      time = str_c(0:23, "00", sep = ":") %>% str_pad(5, "left", "0"),
      format = "netcdf"
    )
  )
  #query to get the ncdf
  server$retrieve("reanalysis-era5-single-levels",
                  query,
                  paste0("era5_ssrd_", y, "_left.nc"))

  query <- r_to_py(
    list(
      variable = "surface_solar_radiation_downwards",
      product_type = "reanalysis",
      area = "75/-180/47/-142", # North, West, South, East
      year = y,
      month = str_pad(2:7, 2, "left", "0"),
      day = str_pad(1:31, 2, "left", "0"),
      time = str_c(0:23, "00", sep = ":") %>% str_pad(5, "left", "0"),
      format = "netcdf"
    )
  )
  #query to get the ncdf
  server$retrieve("reanalysis-era5-single-levels",
                  query,
                  paste0("era5_ssrd_", y, "_right.nc"))

}

years <- as.character(2005:2021)
for(i in 1:length(years)) {
  get_era5(years[i])
}

To explore performance of our solar radiation parameter within a haul-out model
we replace the various Fourier series parameters in our model from the manuscript
with the ERA5 _surface solar radiation downwards_ (`era_ssrd`) parameter. As
with other reanalysis values (from NARR) in the manuscript, the `era-ssrd`
values are matched in time and space to the seal haul-out observation data. We
use the full hourly temporal resolution from ERA5. The glmmLDTS framework used
in the paper does not allow for model comparisons with AIC because of
the reliance on pseudo-likelihood. The `bam` function within the `mgcv` package
provides a quick model fitting option that also allows us to do some model
comparison with AIC. This is sufficient for the general demonstration and
exploartion purposes here
but future research should consider a range of model fitting frameworks and
approaches that might be appropriate.

```{r load-model-data, include=FALSE}
library(targets)
tar_load(spotted_model_data,store = here::here("_targets"))
tar_load(ribbon_model_data,store = here::here("_targets"))
tar_load(bearded_model_data, store = here::here("_targets"))
```

The model specification below was used to specify an `mgcv::bam()` model that
matches the formula used in the manuscript for spotted seals. The `s(speno, bs = "re")` 
term is the smooth term for the random effect. All other predictors are
the same.

```{r bam-m1-spotted, include=TRUE, echo = TRUE}
m1_spotted <- mgcv::bam(
  dry ~ age_sex + s(speno, bs = "re") + 
    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
    poly(day, 3, raw=TRUE) + 
    sin1:poly(day, 3, raw=TRUE) +
    cos1:poly(day, 3, raw=TRUE) +
    sin2:poly(day, 3, raw=TRUE) +
    cos2:poly(day, 3, raw=TRUE) +
    sin3:poly(day, 3, raw=TRUE) +
    cos3:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip + 
    age_sex:poly(day, 4, raw=TRUE),
  data = spotted_model_data,
  family = binomial,
  discrete = TRUE)
```

```{r lag1-spotted, include=FALSE}
lag1_spotted <- acf(resid(m1_spotted), plot=FALSE)[1][[1]]
```

Note, the specification for _m1_spotted_ here does not include
any AR1 structure for temporal autocorrelation. To include this, we need to
provide a value for $\rho$ (or _rho_). We can examine the autocorrelation within
the model and use the lag-1 value for $\rho$ .The value for lag-1 
autocorrelation is `r formatC(as.numeric(lag1_spotted))` which is
rather high but not surprising. We can, now, update our model specification with
a value for $\rho$ as well as the `A1.start` argument with specifies (TRUE/FALSE) the
start point of each block.

```{r bam-m2-spotted, include=TRUE, echo = TRUE}
m2_spotted <- mgcv::bam(
  dry ~ age_sex + s(speno, bs = "re") + 
    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
    poly(day, 3, raw=TRUE) + 
    sin1:poly(day, 3, raw=TRUE) +
    cos1:poly(day, 3, raw=TRUE) +
    sin2:poly(day, 3, raw=TRUE) +
    cos2:poly(day, 3, raw=TRUE) +
    sin3:poly(day, 3, raw=TRUE) +
    cos3:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip + 
    age_sex:poly(day, 3, raw=TRUE),
  data = spotted_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_spotted,
  discrete = TRUE)
```

The model specification for exploring the use of solar radiation is
specified similarly but without all of the Fourier series parameters and
interactions.

```{r bam-ssrd-spotted, include=TRUE, echo = TRUE}
m2_ssrd_spotted <- mgcv::bam(
  dry ~ age_sex + s(speno, bs = "re") + 
    era5_ssrd +
    poly(day, 3, raw=TRUE) + 
    era5_ssrd:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip + 
    age_sex:poly(day, 3, raw=TRUE),
  data = spotted_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_spotted,
  discrete = TRUE)
```

Lastly, the two models were compared with AIC to evaluate whether the
reduction in degrees of freedom with fewer terms in the solar radiation
model was matched with improved explanatory power in the model fit. While the
model and code specified above is for spotted seals, the same approach was
repeated for bearded and ribbon seals.

```{r,bam-ssrd-ribbon-bearded}
m1_ribbon <- mgcv::bam(
  dry ~ age_sex + s(speno, bs = "re") + 
    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
    poly(day, 3, raw=TRUE) + 
    sin1:poly(day, 3, raw=TRUE) +
    cos1:poly(day, 3, raw=TRUE) +
    sin2:poly(day, 3, raw=TRUE) +
    cos2:poly(day, 3, raw=TRUE) +
    sin3:poly(day, 3, raw=TRUE) +
    cos3:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip + 
    age_sex:poly(day, 4, raw=TRUE),
  data = ribbon_model_data,
  family = binomial,
  discrete = TRUE)

lag1_ribbon <- acf(resid(m1_ribbon), plot=FALSE)[1][[1]]

m2_ribbon <- mgcv::bam(
  dry ~ age_sex + s(speno, bs = "re") + 
    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
    poly(day, 3, raw=TRUE) + 
    sin1:poly(day, 3, raw=TRUE) +
    cos1:poly(day, 3, raw=TRUE) +
    sin2:poly(day, 3, raw=TRUE) +
    cos2:poly(day, 3, raw=TRUE) +
    sin3:poly(day, 3, raw=TRUE) +
    cos3:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip + 
    age_sex:poly(day, 3, raw=TRUE),
  data = ribbon_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_ribbon,
  discrete = TRUE)

m2_ssrd_ribbon <- mgcv::bam(
  dry ~ age_sex + s(speno, bs = "re") + 
    era5_ssrd +
    poly(day, 3, raw=TRUE) + 
    era5_ssrd:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip + 
    age_sex:poly(day, 3, raw=TRUE),
  data = ribbon_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_ribbon,
  discrete = TRUE)

m1_bearded <- mgcv::bam(
  dry ~ s(speno, bs = "re") + 
    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
    poly(day, 3, raw=TRUE) + 
    sin1:poly(day, 3, raw=TRUE) +
    cos1:poly(day, 3, raw=TRUE) +
    sin2:poly(day, 3, raw=TRUE) +
    cos2:poly(day, 3, raw=TRUE) +
    sin3:poly(day, 3, raw=TRUE) +
    cos3:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip,
  data = bearded_model_data,
  family = binomial,
  discrete = TRUE)

lag1_bearded <- acf(resid(m1_bearded), plot=FALSE)[1][[1]]

m2_bearded <- mgcv::bam(
  dry ~ s(speno, bs = "re") + 
    sin1 + cos1 + sin2 + cos2 + sin3 + cos3 + 
    poly(day, 3, raw=TRUE) + 
    sin1:poly(day, 3, raw=TRUE) +
    cos1:poly(day, 3, raw=TRUE) +
    sin2:poly(day, 3, raw=TRUE) +
    cos2:poly(day, 3, raw=TRUE) +
    sin3:poly(day, 3, raw=TRUE) +
    cos3:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip,
  data = ribbon_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_bearded,
  discrete = TRUE)

m2_ssrd_bearded <- mgcv::bam(
  dry ~ s(speno, bs = "re") + 
    era5_ssrd +
    poly(day, 3, raw=TRUE) + 
    era5_ssrd:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip,
  data = bearded_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_bearded,
  discrete = TRUE)
```

## Results

To evaluate whether the solar radiation parameter matches our expectations and
compares well with hour of the day, we can visualize the variability of the 
`era5_ssrd` values within
our study area as they relate to hour of the day. The unimodal distribution
is centered around the middle of the solar day with peak solar radiation
coinciding with 13:00 local solar. This suggests solar radiation could be an
informative covariate for capturing unimodal diel patterns in haul-out
behavior.

```{r, include=FALSE}
ssrd_dataset <- readRDS(file = here::here('data/ssrd_dataset.rds'))

pt_ul = st_point(c(160, 70))
pt_br = st_point(c(179, 60))
poly_left <- st_bbox(st_sfc(pt_ul, pt_br,crs = 4326)) |> st_as_sfc()

pt_ul = st_point(c(-160, 70))
pt_br = st_point(c(-179, 60))
poly_right <- st_bbox(st_sfc(pt_ul, pt_br,crs = 4326)) |> st_as_sfc()

# how many random points do we want to generate
n_pts <- 5000

pts1 <- st_sample(poly_left,size = n_pts/2)
pts2 <- st_sample(poly_right, size = n_pts/2)

pts <- c(pts1,pts2)

# Set the seed for reproducibility
set.seed(123)

# Generate random hourly timestamps
get_random_times <- function(start_t, end_t, months, n) {
  tibble(
  time = seq(as.POSIXct(start_t, tz = "UTC"), as.POSIXct(end_t, tz = "UTC"), by = "hours")
)  |> 
  filter(month(time) %in% c(months))  |> 
    filter(month(time) != 2 & mday(time) != 1) |> 
  sample_n(n) %>%
  pull(time)
}

random_times <- get_random_times(start_t = "2005-02-01", end_t = "2021-07-30",
                                 months = c(2,3,4,5,6,7), n = n_pts)

# create a dataset of points with those random times
pts_sf <- st_sf(time=random_times,pts)
pts_vec <- terra::vect(pts_sf)

# list to store for-loop results
pts_ssrd_list <- vector(mode = "list", length = length(years))

years <- 2005:2021

for (i in 1:length(years)) {
  r <- terra::subset(ssrd_dataset, lubridate::year(time(ssrd_dataset)) == years[i])
  v <- terra::subset(pts_vec, lubridate::year(pts_vec$time) == years[i])
  tm <- match(v$time,time(r))
  pts_ssrd_list[[i]] <- terra::extract(r,v,layer=tm, bind=TRUE)
}

library(solaR)

pts_ssrd_sf <- vect(pts_ssrd_list) |> st_as_sf()

pts_ssrd_sf <- pts_ssrd_sf |> 
  mutate(lon = st_coordinates(geometry)[,1]) |> 
  mutate(solar_hour = lubridate::hour(solaR::local2Solar(time,lon)))
```

```{r era5-ssrd-plot, fig.asp = 0.7, include=TRUE, fig.cap="Downward surface solar radiation estimates from the ERA5 climate reanalysis for 5000 random points within the study area between 2005 and 2021. Solar radiation values are presented in Watts per meter^2 and the smoothed line highlights the strong diel pattern."}
ssrd_solarhour_plot <- ggplot(data = pts_ssrd_sf, aes(x=solar_hour, y=value)) + 
  geom_point(size = 0.5, alpha = 0.2, color = "#edd9a3") + 
  geom_smooth(se=TRUE, linewidth = 0.2, color = "#4b2991") +
  # geom_vline(xintercept = 13, linetype="dashed", 
  #               linewidth = 1) +
  scale_x_continuous(breaks = c(3,6,9,12,15,18,21),
                     labels = c("03:00","06:00","09:00","12:00","15:00","18:00",
                                "21:00")) +
  xlab("Local Solar Hour") +
  ylab("Solar Radiation") +
  labs(title = "Diel Pattern of Solar Radiation Values from ERA5 Reanalysis",
       subtitle = "A random sample of hourly ERA5 downward surface solar radiation values for random points within the study area confirms the expected unimodal relationship with local solar hour.",
       caption = "solar radiation values are presented in Watts per meter^2")

ssrd_solarhour_plot
```

The spotted seal model matching the specification from the manuscript resulted in 
`r AIC(m2_spotted,m2_ssrd_spotted)['m2_spotted','df']` degrees of freedom and an AIC value of
`r AIC(m2_spotted,m2_ssrd_spotted)['m2_spotted','AIC']`. The model with solar radiation resulted in 
`r AIC(m2_spotted,m2_ssrd_spotted)['m2_ssrd_spotted','df']` degrees of freedom and an AIC value of
`r AIC(m2_spotted,m2_ssrd_spotted)['m2_ssrd_spotted','AIC']`. Despite the additional terms, the
model with the Fourier series representation of hour of day resulted in a 
lower AIC value and was still preferred for spotted seals. The ribbon seal model matching the specification from the manuscript resulted in 
`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','df']` degrees of freedom and an AIC value of
`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ribbon','AIC']`. The model with solar radiation resulted in 
`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','df']` degrees of freedom and an AIC value of
`r AIC(m2_ribbon,m2_ssrd_ribbon)['m2_ssrd_ribbon','AIC']`. Despite the additional terms, the
model with the Fourier series representation of hour of day resulted in a 
lower AIC value and was still preferred for bearded seals. The bearded seal model matching the specification from the manuscript resulted in 
`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','df']` degrees of freedom and an AIC value of
`r AIC(m2_bearded,m2_ssrd_bearded)['m2_bearded','AIC']`. The model with solar radiation resulted in 
`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','df']` degrees of freedom and an AIC value of
`r AIC(m2_bearded,m2_ssrd_bearded)['m2_ssrd_bearded','AIC']`. Despite the additional terms, the
model with the Fourier series representation of hour of day resulted in a 
lower AIC value and was still preferred for bearded seals.
